\bfseries\LARGE\singlespacing{Estimating \emph{Mycobacterium} tuberculosis transmission in a South African clinic: Spatiotemporal model based on person movements}

### 2025-02-01 doc/paper/manuscript-revised-clean.tex
Updates from Overleaf

\begin{information}

\noindent\textbf{S2 Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic and moves through the waiting room during one morning.

\clearpage

@@

\noindent In crowded indoor settings in high tuberculosis burden countries, there is a high risk of tuberculosis transmission. We developed a model to study how tuberculosis is spread in a South African clinic during the COVID-19 pandemic when masks were mandatory and ventilation was improved. We tracked the movements of clinic attendees with video sensors and matched them with clinical records to find where and when infectious tuberculosis patients were present. We used this information to estimate the concentration of infectious particles in the air and the risk of infection for individuals. Over five days, the sensors tracked nearly 1,500 people in the clinic. Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room. The average risk of infection per person was found to be \chdeleted{0.1\%}\chadded{0.05\%}. The risk increased with more close contact with infectious patients and longer time spent in the clinic. Without masks, the risk was 1.3 times higher. With lower ventilation rates, similar to pre-pandemic levels, the risk was \chdeleted{2.2}\chadded{2.1} times higher. This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.

\end{information}

\begin{abstract}\normalfont
The risk of \emph{Mycobacterium tuberculosis (Mtb)} transmission can be high in crowded clinics. We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation. We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when \chadded{community or surgical} masks were compulsory and ventilation was increased. We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients. We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection. Over five days, video sensors tracked 1,438 clinic attendees. CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room. The estimated risk of infection per clinic attendee was \chdeleted{0.1\% (80\%-credible interval (CrI) 0.01\%$-$0.1\%)}\chadded{0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%)}. It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and \chdeleted{2.2-fold (95\%-CrI 0.9$-$5.2)}\chadded{2.1-fold (95\%-CrI 0.9$-$5.0)} higher with pre-pandemic ventilation rates\chadded{, emphasizing the importance of ventilation}. Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.

%TC:ignore

\vspace{2em}

%\noindent\textbf{Word count:} \abstractcount{manuscript}words (max. 500)

\vspace{0.5em}

\noindent\textbf{Keywords:} Mycobacterium tuberculosis, airborne infection, person tracking, spatiotemporal model
% maximum of 3-5 keywords
\newpage

\sloppy
\raggedbottom
%TC:endignore

@@

\emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, is transmitted via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}. \emph{Mtb} is primarily carried in small particles of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}. Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}. Healthcare facilities are high-risk settings \chdeleted{because infectious individuals are more likely to visit clinics}\chadded{where infectious individuals are more likely to visit}\cite{McCreesh2020IJTLD}. A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.

@@

Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa. The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm. It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}. We defined three areas within the clinic (\Cref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m). The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB. \chadded{The clinic was naturally ventilated by opening all doors and windows throughout the day.}

@@

We collected environmental and clinical data and data on the movements of patients in the clinic. Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (\Cref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals. On October~13, the corridor's CO$_2$ \chdeleted{levels}\chadded{measurements} were missing due to a malfunctioning device and the waiting room's CO$_2$ \chdeleted{levels}\chadded{measurements} were used for the corridor because their CO$_2$ \chdeleted{levels}\chadded{measurements} were similar on the other days \chadded{(see time-varying CO$_2$ levels and number of persons per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives})}.

@@

We used an anonymous person tracking system using video sensors (Xovis, Zollikofen, Switzerland) to track the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (\Cref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person. To address this, we created an R Shiny application (Fig.~\zref{fig:shiny-app} in \supp) to rejoin interrupted tracks as described in Section~\zref{sec:setting-and-data} in \supp. We also labeled the tracks to identify the movements of clinic staff. \Cref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.

@@

Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of \emph{Mtb} infection $P$ as
\begin{align}
    P = \frac{C}{S} = 1 - \exp\left(-\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate \chadded{per person} (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$). The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection. The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.

We extended the Wells-Riley model, allowing spatial variation in the quanta concentration by combining patient-tracking data with clinical records. This way, the spatiotemporal location of infectious patients who generate quanta could be determined. A detailed description of our extension is provided in Section~\zref{sec:spattemp-model} in \supp, together with an illustrative example using simulated data (Fig.~\zref{fig:toy-example} in \supp). In the following, we describe the six main components of our modeling approach (\Cref{fig:modeling-flow}).

@@

(2)~In addition to diagnosed TB patients, we also assumed a number of undiagnosed (subclinical) TB patients. The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}. Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig.~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients. Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious. We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig.~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}. Since mask use \chadded{(community or surgical masks)} was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.

@@

(3)~We calculated \chadded{time-varying} room occupancy from the video sensor data and combined it with the CO$_2$ measurements \chadded{(Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp)} to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.

@@

(5)~Quanta removal occurs through the replacement of contaminated indoor air with outdoor air, inactivation of bacteria in airborne particles, and gravitational settling. The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate. We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig.~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,\chdeleted{quanta }h$^{-1}$ (95\%-CrI 0.1\,\chdeleted{quanta }h$^{-1}$\,$-$\,7.1\,\chdeleted{quanta }h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}. \chdeleted{We set the gravitational settling rate to zero and ignore sedimentation because} Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of \chdeleted{just} $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, \chadded{which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-4}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-4}$m\,s$^{-1}$. Considering a drop height of 1.7\,m \chadded{(average height of visitors)}, the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).}

@@

(6)~\chdeleted{Based on the quanta generation, diffusion, and removal process, the quanta concentration $N$ (quanta\,m$^{-3}$) at time $t$ was computed as}\chadded{The quanta concentration was computed separately for the waiting room, corridor, and TB room. Each room was discretized into a grid of cubic cells, each cell covering a squared area of 0.0625\,m$^2$. We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as} 
\chadded{
\begin{align}\label{eq:spattemp-N}
    % \chdeleted{\underbrace{N_{t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{t-1}}_{\text{prev. conc.}} + \underbrace{I_t \cdot q \cdot 1\text{s}}_{\text{generation}}) \cdot 1\text{s}\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_t + \lambda\cdot 1\text{s}\right)}_{\text{removal}} ~.}
     \underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q\cdot1\,s}_{\text{generation}}\cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k)\cdot1\,s\right)}_{\text{removal}} ~.
\end{align}
}
where \chdeleted{$D$}\chadded{$D_t$} is the diffusion constant (m$^2$\,s$^{-1}$) \chadded{varying by daytime}, $\Delta$ is the Laplace operator (second-order differential operator), \chdeleted{$I_t$ is the number of infectious individuals in space, } \chadded{$I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$)}, $q$ is the quantum generation rate (quanta\,s$^{-1}$), \chdeleted{$AER$ is the}\chadded{$AER_{r,t}$ is the room-specific} air change rate \chadded{varying by daytime}(\chdeleted{quanta}s$^{-1}$), $\lambda$ is the bacterial inactivation rate (\chdeleted{quanta}s$^{-1}$), \chadded{and $k$ is the gravitational settling rate (s$^{-1}$)}. \chadded{We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.}

The patient-specific risk of infection depends on the cumulative exposure to infectious quanta and was computed using the Wells-Riley equation as
\begin{align}
    % \chdeleted{P = 1-\exp\left(-\sum_c \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p_a \cdot 1s\right),}
    \chadded{P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),}
\end{align}
where \chdeleted{$N_{c,t}$ is the quanta concentration in cell space $c$ at time $t$, $\mathbb{I}_{c,t}$ indicates whether the person was in $c$ at $t$}\chadded{$\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$}, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$. We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$. \chadded{In line with a previous modeling study\cite{McCreesh2021BMJGlobalHealth}, we assumed that the mandatory use of community and surgical masks during our study had only effects on the exhalation of \emph{Mtb} particles and no effect on \emph{Mtb} infection of the person wearing the mask.}

\subsection*{\chdeleted{Monte Carlo simulation}}

\chdeleted{Simulations are described in detail in Section~\zref{sec:estimation} in \supp. Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and the TB room. Each room was rasterized into a grid of cubic cells with an area of 0.25$^2$\,m$^2$. The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second. We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters. In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.}

@@

Data were collected during the COVID-19 pandemic when infection control measures were in place. To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask-wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie assuming an air change rate of 6~air changes per hour, corresponding to the average rate in the pilot study in 2019 in the same clinic\cite{Zurcher2022JID}. Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}. \chdeleted{Except for scenario-specific modifications, all other modeling assumptions and parameters were identical in each Monte Carlo simulation.}

@@

We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics. \chadded{We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp). In each simulation, we computed the individual risk of infection based on the cumulative quanta exposure during the clinic visit. We computed the average quanta concentration by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the mean risk of infection per clinic attendee across simulations. The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI).} \chdeleted{Spatiotemporal quanta concentration and risk of infection per clinic attendee were estimated based on 5,000 Monte Carlo simulations. The average quanta concentration was computed by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the risk of infection was summarized with the mean and 80\%-credible interval (CrI).} The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%). We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modeled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min). Analyses were performed in R (version 4.3.1) and Stan (version 2.26.1). Additional results not reported in the manuscript are shown in Section~\zref{sec:additional-results} in \supp.

@@

%TC:ignore
\section*{Acknowledgements}
We are grateful to the clinical manager and the staff members at the primary care clinic. We also would like to thank the City of Cape Town, South Africa for the use of one of their clinic facilities and their support. Our research findings and recommendations do not represent an official view of the city of Cape Town.

>>>>

\begin{abstract}\normalfont
The risk of \emph{Mycobacterium tuberculosis (Mtb)} transmission can be high in crowded clinics. We developed a spatiotemporal model of airborne \emph{Mtb} transmission based on the Wells-Riley equation. We collected environmental, clinical and person-tracking data in a South African clinic during COVID-19, when community or surgical masks were compulsory and ventilation was increased. We matched person movements with clinical records to identify the spatiotemporal location of infectious TB patients. We modeled the concentration of infectious doses (quanta) and estimated the individual risk of infection. Over five days, video sensors tracked 1,438 clinic attendees. CO$_2$ levels were low (median 431\,ppm, IQR 406\,ppm$-$458\,ppm); the quanta concentration was higher in the morning than in the afternoon, and highest in the waiting room. The estimated risk of infection per clinic attendee was 0.05\% (80\%-credible interval (CrI) 0.01\%$-$0.06\%). It increased with the number of close contacts with infectious patients and the time spent in the clinic, and was 1.3-fold (95\%-CrI 1.2$-$1.4) higher in scenarios without mask use and 2.1-fold (95\%-CrI 0.9$-$5.0) higher with pre-pandemic ventilation rates, emphasizing the importance of ventilation. Spatiotemporal modeling can identify high-risk areas and evaluate the impact of infection control measures in clinics.

\sloppy
\raggedbottom

@@

\noindent In crowded indoor settings in high tuberculosis burden countries, there is a high risk of tuberculosis transmission. We developed a model to study how tuberculosis is spread in a South African clinic during the COVID-19 pandemic when masks were mandatory and ventilation was improved. We tracked the movements of clinic attendees with video sensors and matched them with clinical records to find where and when infectious tuberculosis patients were present. We used this information to estimate the concentration of infectious particles in the air and the risk of infection for individuals. Over five days, the sensors tracked nearly 1,500 people in the clinic. Air quality was generally good, but the concentration of infectious particles was higher in the morning and peaked in the waiting room. The average risk of infection per person was found to be 0.05\%. The risk increased with more close contact with infectious patients and longer time spent in the clinic. Without masks, the risk was 1.3 times higher. With lower ventilation rates, similar to pre-pandemic levels, the risk was 2.1 times higher. This modeling can help identify high-risk areas for tuberculosis transmission and evaluate infection control measures in healthcare facilities.

@@

\emph{Mycobacterium tuberculosis} (\emph{Mtb}), the causative agent of TB, is transmitted via respiratory particles in the exhaled air of infectious persons\cite{Rieder1999,Patterson2021Tuberculosis}. \emph{Mtb} is primarily carried in small particles of $1-7\mu$m\cite{Fennelly2020Lancet}, which can survive in the air for multiple hours\cite{Loudon1969AMRRD}. Survival of airborne pathogens is greater in crowded, poorly ventilated indoor environments\cite{Rieder1999,CPS2013Book,Nardell1991ARRD,Wang2021Science,Morawska2021}. Healthcare facilities are high-risk settings where infectious individuals are more likely to visit\cite{McCreesh2020IJTLD}. A modeling study in South Africa estimated that 4\% to 14\% of adult TB cases originate from primary care clinics\cite{McCreesh2022BMJGlobalHealth}.

@@

Based on a pilot study from 2019 in the same clinic\cite{Zurcher2022JID,Zurcher2020BMJ}, we collected environmental data (indoor CO$_2$ levels), clinical data (TB status), and video sensor data (person movements) for five days during the COVID-19 pandemic in October/November 2021 (October 13, 15, 25, and November 4, 5) at a primary care clinic in Cape Town, South Africa. The clinic offers both TB and HIV services and other basic clinical services, Monday through Friday, from 8\,am to 4\,pm. It is located in a large settlement of formal and semi-formal housing where TB and HIV are highly prevalent\cite{Wood2007AMJRCCD,Middelkoop2011JAIDS}. We defined three areas within the clinic (Fig~\ref{fig:floor-plan}): the waiting room (10.55m $\times$ 5.50m $\times$ 3.00m), corridor (12.45m $\times$ 2.20m $\times$ 2.50m), and TB room (4.75m $\times$ 3.50m $\times$ 3.00m). The TB room was mainly used for the treatment of diagnosed TB patients but also for the screening of suspected patients with respiratory symptoms who were ultimately not diagnosed with TB. The clinic was naturally ventilated by opening all doors and windows throughout the day.

@@

We collected environmental and clinical data and data on the movements of patients in the clinic. Four devices monitored indoor CO$_2$ levels (CO$_2$ Logger CL11, Rotronic, Bassersdorf, Switzerland) in the waiting room, corridor, and TB room (Fig \ref{fig:floor-plan}), in parts per million (ppm) at 1-min intervals. On October~13, the corridor's CO$_2$ measurements were missing due to a malfunctioning device and the waiting room's CO$_2$ measurements were used for the corridor because their CO$_2$ measurements were similar on the other days (see time-varying CO$_2$ levels and number of persons per room by clinic day in Fig.~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp, Section \zref{sec:add-descriptives}).

@@

We used an anonymous person tracking system using video sensors (Xovis, Zollikofen, Switzerland) to track the movements of people (clinic staff, patients, and other visitors) throughout the clinic at 1-second intervals (Fig~\ref{fig:floor-plan}). The resulting time-stamped tracking data consisted of a person’s height, their position recorded as x-y coordinates, and a unique ID for each person's track while in the clinic. People in the clinic could contribute multiple tracks if they moved out of the range of a sensor or were not recognized as a person. To address this, we created an R Shiny application (Fig~\zref{fig:shiny-app} in \supp) to rejoin interrupted tracks as described in Section~\zref{sec:setting-and-data} in \supp. We also labeled the tracks to identify the movements of clinic staff. Fig~\ref{fig:tracking-examples} shows four examples of person movements after rejoining interrupted tracks.

@@

Our spatiotemporal model is based on the Wells-Riley equation\cite{Riley1961Book}, which estimates the risk (probability in \%) of \emph{Mtb} infection $P$ as
\begin{align}
    P = \frac{C}{S} = 1 - \exp\left(-\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate per person (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (h), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$). The probability of infection is estimated with a Poisson relation, considering the stochastic behavior of airborne infection, where one quantum corresponds to a 67\% risk of infection. The main assumption of the Wells-Riley model is a well-mixed airspace, which means that the quanta concentration $N = Iq/Q$ (quanta m$^{-3}$) is the same everywhere in the room.

We extended the Wells-Riley model, allowing spatial variation in the quanta concentration by combining patient-tracking data with clinical records. This way, the spatiotemporal location of infectious patients who generate quanta could be determined. A detailed description of our extension is provided in Section~\zref{sec:spattemp-model} in \supp, together with an illustrative example using simulated data (Fig~\zref{fig:toy-example} in \supp). In the following, we describe the six main components of our modeling approach (Fig~\ref{fig:modeling-flow}).

@@

(2)~In addition to diagnosed TB patients, we also assumed a number of undiagnosed (subclinical) TB patients. The number of undiagnosed TB patients may be proportional to the number of diagnosed TB patients \cite{Berhanu2023CID,Moyo2022LancetID,Patterson2024PNAS}. Therefore, we modeled the number of undiagnosed TB patients with a multinomial distribution based on the counts of the number of diagnosed patients in the clinical data in October/November 2021 (Fig~\zref{fig:undiagnosed-distribution} in \supp): 6 days (12\%) with zero, 12 (41\%) with one, 8 (28\%) with two, 2 (7\%) with three, and 1 (3\%) with four diagnosed TB patients. Undiagnosed patients were randomly sampled among all clinic attendees with the same probability, excluding clinic staff who were not considered infectious. We further assumed that diagnosed and undiagnosed TB patients generated quantum at the same rate $q$, modeled using a lognormal prior distribution considering two different activity levels (Fig~\zref{fig:quanta-distribution} in \supp): median of 1.08\,quanta h$^{-1}$ (95\%-credible interval [CrI] 0.003\,quanta h$^{-1}$\,$-$\,386\,quanta h$^{-1}$) while sitting and 2.81\,quanta h$^{-1}$ (95\%-CrI 0.008\,quanta h$^{-1}$\,$-$\,987\,quanta h$^{-1}$) while walking\cite{Mikszewski2021GF,Buonanno2020EI,Banholzer2024PGPH}. Since mask use (community or surgical masks) was mandatory for all clinic staff and attendees during the study, we assumed a mean reduction in $q$ by 75\% (95\%-CrI 56\%\,$-$\,86\%)\cite{Dharmadhikari2012AJRCCM,McCreesh2021BMJGlobalHealth}.

@@

(3)~We calculated time-varying room occupancy from the video sensor data and combined it with the CO$_2$ measurements (Figs~\zref{fig:co2-levels}-\zref{fig:people-over-time-by-day} in \supp) to the estimate air change rate during the morning (8\,am to 12\,am) and afternoon (12\,am to 4\,pm) of each clinic day using a transient mass balance model\cite{Batterman2017IJERPH}.

@@

(5)~Quanta removal occurs through the replacement of contaminated indoor air with outdoor air, inactivation of bacteria in airborne particles, and gravitational settling. The removal rate is thus the sum of the air change rate, the inactivation rate of \emph{Mtb}, and the gravitational settling rate. We modeled the bacterial inactivation rate of \emph{Mtb} using a lognormal prior distribution (Fig~\zref{fig:undiagnosed-distribution} in \supp) with a median of 1\,h$^{-1}$ (95\%-CrI 0.1\, h$^{-1}$\,$-$\,7.1\,h$^{-1}$)\cite{Loudon1969AMRRD,Lever2000LettersAppliedMicrobio,Gannon2007ResVetSci,Klein2014IJMyco}. Particles with \emph{Mtb} in the size range of $1-7\mu$m\cite{Fennelly2020Lancet} have a settling velocity of $3.5\cdot 10^{-5}-1.5\cdot 10^{-3}$m\,s$^{-1}$ in still air\cite{Vuorinen2020SafSci}, which we modeled using a Gamma distribution with mean $7.7\cdot 10^{-4}$m\,s$^{-1}$ and standard deviation $3.7\cdot 10^{-4}$m\,s$^{-1}$. Considering a drop height of 1.7\,m (average height of visitors), the median gravitational settling rate was assumed to be 1.5 h$^{-1}$ (95\%-CrI 0.5 h$^{-1}$ – 3.5 h$^{-1}$).

@@

(6)~The quanta concentration was computed separately for the waiting room, corridor, and TB room. Each room was discretized into a grid of cubic cells, each cell covering a squared area of 0.0625\,m$^2$. We assumed that the quanta concentration was vertically well-mixed and computed the quanta concentration $N_{r,c,t}$ in room $r$ and cell $c$ at time $t$ as
\begin{align}\label{eq:spattemp-N}
     \underbrace{N_{r,c,t}}_{\text{new conc.}} = \underbrace{\left(D \Delta (\underbrace{N_{r,c,t-1}}_{\text{old conc.}} + \underbrace{I_{r,c,t} \cdot q\cdot1\,s}_{\text{generation}}\cdot V_{r,c}^{-1})\right)}_{\text{diffusion}} \cdot \underbrace{\exp\left(-(AER_{r,t} + \lambda + k)\cdot1\,s\right)}_{\text{removal}} ~.
\end{align}
where $D_t$ is the diffusion constant (m$^2$\,s$^{-1}$) varying by daytime, $\Delta$ is the Laplace operator (second-order differential operator), $I_{r,c,t}$ is the number of infectious individuals, $V_{r,c}$ is the room-specific cell volume (m$^3$), $q$ is the quantum generation rate (quanta\,s$^{-1}$), $AER_{r,t}$ is the room-specific air change rate varying by daytime(s$^{-1}$), $\lambda$ is the bacterial inactivation rate (s$^{-1}$), and $k$ is the gravitational settling rate (s$^{-1}$). We modeled the spatiotemporal quanta concentration during clinic hours from 8\,am to 4\,pm and updated it every second, corresponding to the frequency of the person-tracking data.

The patient-specific risk of infection depends on the cumulative exposure to infectious quanta and was computed using the Wells-Riley equation as
\begin{align}
    P = 1-\exp\left(-\sum_r \sum_c \sum_t N_{r,c,t} \cdot \mathbb{I}_{r,c,t} \cdot p_a\cdot1\,s\right),
\end{align}
where $\mathbb{I}_{r,c,t}$ indicates whether the person was in cell $c$ of room $r$ at time $t$, and $p_a$ (m$^3$ s$^{-1}$) is the breathing rate for activity level $a$. We distinguished between walking ($p_\mathrm{walk}$ = 1.33\,m$^3$\,h$^{-1}$) and sitting activities ($p_\mathrm{sit}$ = 0.51\,m$^3$\,h$^{-1}$)\cite{Adams1993}, based on whether the the person had moved by more or less than 0.25\,m\,s$^{-1}$. In line with a previous modeling study\cite{McCreesh2021BMJGlobalHealth}, we assumed that the mandatory use of community and surgical masks during our study had only effects on the exhalation of \emph{Mtb} particles and no effect on \emph{Mtb} infection of the person wearing the mask.

\subsection*{Monte Carlo simulation}

Simulations are described in detail in Section~\zref{sec:estimation} in \supp. Briefly, we simulated the spatiotemporal quanta concentration separately for the clinic's waiting room, corridor, and the TB room. Each room was rasterized into a grid of cubic cells with an area of 0.25$^2$\,m$^2$. The quanta concentration was modeled during clinic hours from 8\,am to 4\,pm and updated every second. We used a Monte-Carlo simulation approach to consider uncertainty in modeling parameters. In each simulation, we sampled uncertain modeling parameters from their prior distributions and computed the spatiotemporal quanta concentration and the risk of infection per clinic attendee.

@@

Data were collected during the COVID-19 pandemic when infection control measures were in place. To model their effectiveness, we considered two hypothetical scenarios: 1)~no mask-wearing, \ie assuming no reduction in the quantum generation rate; and 2)~pre-pandemic ventilation conditions, \ie assuming an air change rate of 6~air changes per hour, corresponding to the average rate in the pilot study in 2019 in the same clinic\cite{Zurcher2022JID}. Furthermore, we examined the impact of different assumptions about the number of infectious patients in the clinic: 1)~we considered patients clinically suspected of having TB as infectious because these patients may generate aerosolized \emph{Mtb} even when not diagnosed with TB at their first visit\cite{Patterson2024PNAS}; 2)~we randomly selected infectious patients among all clinic attendees in proportion to the prevalence of TB in the South African population\cite{Moyo2022LancetID}.

@@

We summarized clinical (diagnosed and suspected TB patients), environmental (time-varying CO$_2$ levels and air change rates), and person-tracking data (spatiotemporal number of tracks, visit time, and close-contact encounters) using descriptive statistics. We modeled the spatiotemporal quanta concentration with 5,000 Monte Carlo simulations, considering uncertainty in modeling parameters (Section~\zref{sec:estimation} in \supp). In each simulation, we computed the individual risk of infection based on the cumulative quanta exposure during the clinic visit. We computed the average quanta concentration by daytime (morning: 8\,am to 12\,am, afternoon: 12\,am to 4\,pm) and the mean risk of infection per clinic attendee across simulations. The scenario-specific infection risk distribution was shown with a histogram and summarized with the mean and 80\%-credible interval (CrI). The mean risk of infection per clinic attendee was categorized as low (<0.1\%), medium (0.1-1\%), and high (>1\%). We used a Bayesian Beta regression model to estimate the odds ratio (OR, posterior mean and 95\%-CrI) of the modeled risk of TB infection in association with the visit time, duration of close contact with other clinic attendees, and frequency of close-contact encounters (proximity to another attendee <1\,m for a duration of >1\,min). Analyses were performed in R (version 4.3.1) and Stan (version 2.26.1). Additional results not reported in the manuscript are shown in Section~\zref{sec:additional-results} in \supp.

@@

\section*{Supplementary Information}

\noindent\textbf{S1 Video:} Animated sequence of the spatiotemporally varying quanta concentration as an infectious patient enters the clinic and moves through the waiting room during one morning.

\section*{Acknowledgments}
We are grateful to the clinical manager and the staff members at the primary care clinic. We also would like to thank the City of Cape Town, South Africa for the use of one of their clinic facilities and their support. Our research findings and recommendations do not represent an official view of the city of Cape Town.
